Categories of Optics

### 2018-07-24 paper.tex
Start on a mixed optics section

\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\const}{\mathrm{const}}
\DeclareMathOperator{\ob}{ob}
\DeclareMathOperator{\copr}{copr}
\newcommand{\inl}{\mathrm{inl}}
\newcommand{\inr}{\mathrm{inr}}
\DeclareMathOperator{\im}{im}
\newcommand{\act}{\cdot}

@@

\begin{proposition}
  There is a functor $\iota : \C \times \C^\op \to \Optic_\otimes$, which on objects is given by $\iota(S, S') = (S, S')$ and on a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \rep{\lambda_A^{-1} f}{g \lambda_{A'}}$.
\end{proposition}
\begin{proof}
  Graphically, this is:
  \begin{center}
    \input{diagrams/iota.tikz}
  \end{center}

@@

\begin{proposition}\label{prop:change-of-action-monoidal}
  For monoidal $F : \C \to \D$, the induced $\Optic(F) : \Optic_\C \to \Optic_\D$ is also monoidal.
\end{proposition}
\begin{proof}
The structure morphisms for monoidality are given by lifting the structure morphisms for $F$:
\begin{align*}
\phi_{(S, S'), (T, T')} &:= \iota(\phi_{S, T}, \phi_{S', T'}) &&: F(S, S') \otimes F(T, T') \hto F((S, S') \otimes (T, T')) \\
\phi &:= \iota(\phi, \phi) &&: (I, I) \hto F(I, I)
\end{align*}
The axioms follow by lifting the axioms for $F$. More difficult is verifying naturality: \todo{todo!}
\end{proof}

@@

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $MA$. The tensor product in $\M$ will also be written as juxtaposition, so $(NM)A \cong N(MA)$.
\end{definition}

@@

The following are analogous to Propositions~\ref{prop:optic-is-cat} and~\ref{prop:change-of-action-monoidal}.

\begin{proposition}
  We have a category $\Optic_\M$ and a functor $\iota : \C \times \C^\op \to \Optic_\M$. \qed
\end{proposition}

\begin{definition}
Given two categories equipped with monoidal actions $(\M, \C)$ and $(\N, D)$, a \emph{morphism of actions} is a monoidal functor $F^\bullet : \M \to N$ and a functor $F : \C \to \D$ that commutes with the actions, in the sense that there exist natural isomorphisms
  \begin{align*}
  \phi &: F(I A) \to (F^\bullet I)(FA) \\
  \phi_{M,A} &: F(MA) \to (F^\bullet M)(F A)
  \end{align*}
satisfying conditions analogous to those for a monoidal functor.
\end{definition}

\begin{proposition}\label{prop:change-of-action}
If $F : (\M, \C) \to (\N, D)$, is a morphism of actions, there is an induced functor $\Optic(F) : \Optic_\M \to \Optic_\N$. \qed
\end{proposition}

@@

\begin{proposition}\label{prop:lawful-lens-laws}
  A concrete lens given by $\fget$ and $\fput$ is lawful (in our sense) iff it obeys the three concrete lens laws.
\end{proposition}
\begin{proof}
  We begin by giving $\Twoptic_\times(S, A)$ the same treatment as $\Optic_\times(S, A)$. Using Yoneda reduction and the universal property of the product, as we did for lenses themselves, it is isomorphic to the set
  \[\conctwice_\times(S, A) := \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)\]
  with the isomorphism given by
  \begin{align*}
    \Phi(\repthree{l}{c}{r}) = (&\pi_2 l, \\
                                   &\pi_2 c (\pi_1 l \times A), \\
                                   &r(\pi_1 c (\pi_1 l \times A) \times A))
  \end{align*}
  Now suppose we are given a lens $p$ that corresponds concretely to $(\fget, \fput)$, so $p = \rep{[\id_S, \fget]}{\fput}$. Evaluating $\outside$ on this gives:
  \begin{align*}
    \outside(\rep{[\id_S, \fget]}{\fput}) = \fput [\id_S, \fget]
  \end{align*}
  so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

@@

\begin{definition}
  Suppose $\C$ has finite coproducts, and let $\C$ act on itself via the bifunctor $\sqcup : \C \times \C \to \C$. The \emph{category of prisms} is the category of lawful optics with respect to $\sqcup$.
\end{definition}

@@

There is a common pattern in many of the examples to follow: for every object $A \in \C$, the evaluation-at-$A$ functor $-A : \M \to \C$ has a right adjoint, say $R_A : \C \to \M$. To fix notation, let \[\overrightarrow{(-)} : \C(F \act A, S) \to \M(F, R_{A} S) : \overleftarrow{(-)}\] denote the homset bijection, so the unit and counit are:
  \begin{align*}
  	\eta_F &: F \to R_A (F \act A) \\
    \eta_F &:= \overrightarrow{\id_{F \act A}} \\
    \varepsilon_{S} &: (R_A S) \cdot A \to S \\
    \varepsilon_{S} &:= \overleftarrow{\id_{R_{A} S}}
  \end{align*}

@@

In this case we can always eliminate the coend and find a concrete description of an optic:
\begin{align*}
&\int^{F \in \M} \C(S, F\act A) \times \C(F\act A', S') \\
&\cong \int^{F \in \M} \C(S, F \act A) \times \M(F, R_{A'} S') \\
&\cong \C(S,  (R_{A'} S') \act A)
\end{align*}
A concrete optic is therefore a map $\funzip : S \to (R_{A'} S') \act A$. The above isomorphism sends $\funzip$ to the element $\rep{\funzip}{\varepsilon_{S'}}$. In the other direction, given $\rep{l}{r}$ we have $\funzip = (\overrightarrow{r} \otimes A)l$.

@@

\begin{definition}
The category $\Traversal$ of traversals is the category of lawful optics for the action of $\Traversable$ on $\Set$ given by evaluation. (Yes, the names $\Traversal$ vs\@. $\Traversable$ are confusing.)
\end{definition}

@@

Suppose $\C$ is a monoidal closed category and $T : \C \to \C$ is a strong monad. We can find a concrete description of the optics for the action $\monact$, using the same method as for linear lenses.

\begin{align*}
\Optic((S, S'), (A, A'))
&= \int^{M \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A', S') \\
&= \int^{M \in \C} \C_T(S, M \monact A) \times \C(M \otimes A', T S') \\
&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M, \homC(A', T S')) \\
&\cong \C_T(S, \homC(A', T S') \monact A) \\
&\cong \C(S, T(\homC(A', T S') \otimes A))
\end{align*}

Concrete effectful lenses therefore consist of a single morphism $\munzip : S \to T(\homC(A', T S') \otimes A)$. The category $\C$ includes into $\C_T$ in a way that preserves the action of $\C$, so there is an induced inclusion $\Optic_\otimes \to \Optic_\monact$.

\begin{proposition}
A concrete effectful lens is lawful iff
  \begin{align*}
    \mu_S T(\ev_{A, TS}) \; \munzip &= \eta_S \\
    T(\eta_{\homC(A, TS) \otimes A}\coev_{\homC(A, TS), A} \otimes A)\munzip &= T((\munzip \circ_T -) \otimes A)\munzip
  \end{align*}
  where \[ \munzip \circ_T - : \homC(A, TS) \to \homC(A, T(\homC(A, TS) \otimes A)) \] denotes internal Kleisli composition and \[\coev_{\homC(A, TS), A} : \homC(A, TS) \to \homC(A, \homC(A, TS) \otimes A) \] is coevaluation.
\end{proposition}
  Or, if you prefer do-notation, the two laws are:
\begin{multicols}{2}
\begin{minted}{haskell}
do (c, a) <- munzip s
   c a
==
return s
\end{minted} 
~\columnbreak
\begin{minted}{haskell}
do (c, a) <- munzip s
   let f a' = do
     s' <- c a'
     munzip s'
   return (f, a)
==
do (c, a) <- munzip s
   let f a' = (c, a')
   return (f, a)
\end{minted}
\end{multicols}

\begin{proof}
This is exactly analogous to Proposition~\ref{prop:concrete-linear-lawful} for linear lenses, with the unit and multiplication of the monad inserted in the appropriate places.
%\begin{align*}
%&\int^{M, N \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, N \monact A) \times \C_T(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, N \monact A) \times \C(N, \homC(A, TS)) \\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C_T(M \monact A, \homC(A, TS) \monact A)\\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M \times A, T(\homC(A, TS) \times A))\\
%&\cong \int^{M \in \C} \C_T(S, M \monact A) \times \C(M, \homC(A, T(\homC(A, TS) \monact A)))\\
%&\cong \C_T(S, \homC(A, T(\homC(A, TS) \times A)) \times A) \\
%&\cong \C(S, T(\homC(A, T(\homC(A, TS) \times A)) \times A))
%\end{align*}
\end{proof}

@@

@@

Working in a cartesian closed category, we can find a concrete description of optics for the action $\monact : \C \times \C_{T_Q} \to \C_{T_Q}$ that is closer to that for ordinary lenses.
\begin{align*}
\Optic((S, S'), (A, A'))
&= \int^{M \in \C} \C_{T_Q}(S, M \monact A) \times \C_{T_Q}(M \monact A', S') \\
&= \int^{M \in \C} \C(S, \homC(Q, M \times A \times Q)) \times \C_{T_Q}(M \monact A', S') \\
&\cong \int^{M \in \C} \C(S \times Q, M \times A\times Q) \times \C_{T_Q}(M \monact A', S') \\
&\cong \int^{M \in \C} \C(S \times Q, M) \times \C(S \times Q, A\times Q) \times \C_{T_Q}(M \monact A', S') \\
&\cong \C(S \times Q, A \times Q) \times \C_{T_Q}((S \times Q) \monact A', S') \\
&\cong \C_{T_Q}(S, A) \times \C_{T_Q}(S \times Q \times A', S')
\end{align*}
By analogy with ordinary lenses, let us call these maps $\mget$ and $\mput$.

@@

\begin{proposition}
\todo{Translate this from code}
A stateful lens given by
\begin{minted}{haskell}
mget :: s -> State q a
mput :: s -> q -> a -> State q s
\end{minted}
is lawful iff: \\
\begin{minipage}{\textwidth}
\begin{multicols}{3}
\begin{minted}{haskell}
do
  q <- getState
  a <- mget s
  mput s q a
==
return s
\end{minted}
~\columnbreak
\begin{minted}{haskell}
do s' <- mput s q a
   mget s'
==
return a
\end{minted}
~\columnbreak
\begin{minted}{haskell}
let (s', q') 
  = runState (mput s q1 a1) q2
in mput s' q' a2
==
mput s q1 a2
\end{minted}
\end{multicols}
\end{minipage}
\end{proposition}
We may as well also name these the $\fget\fput$, $\fput\fget$ and $\fput\fput$ laws.
\begin{proof}
%We similarly calculate
%\begin{align*}
%&\int^{M, N \in \C} \C_{T_Q}(S, M \monact A) \times \C_{T_Q}(M \monact A, N \monact A) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C(S, \homC(Q, M \times A \times Q)) \times \C(M \times A, \homC(Q, N \times A \times Q)) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C(S \times Q, M \times A \times Q) \times \C(M \times A \times Q, N \times A \times Q) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C(S \times Q, M) \times \C(S \times Q,  A \times Q) \times \C(M \times A \times Q, N \times A \times Q) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{N \in \C} \C_{T_Q}(S, A) \times \C(S \times Q \times A \times Q, N \times A \times Q) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{N \in \C} \C_{T_Q}(S, A) \times \C(S \times Q \times A \times Q, N) \times \C(S \times Q \times A \times Q, A \times Q) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \C_{T_Q}(S, A)\times \C_{T_Q}((S \times Q) \monact A, A) \times \C_{T_Q}((S \times Q \times A \times Q) \monact A, S)
%\end{align*}
\todo{This is a mess}
\end{proof}

@@

\subsection{Degenerate Optics}

We finish with some almost-examples of optic variants, the \mintinline{haskell}{Getter}s and \mintinline{haskell}{Review}s of the Haskell library, where an optic $(S, S') \hto (A, A')$ corresponds simply to a morphism $S \to A$ or $A' \to S'$ respectively. These can fit into the coend formalism, with a slight modification.

If we have a monoidal category $\M$ that acts on two different categories $\C_1$ and $\C_2$, we may form the `optic' category:
\begin{align*}
  \Optic_\M((S, S'), (A, A')) := \int^{M \in \M} \C_1(S, MA) \times \C_2(MA', S')
\end{align*}
We lose, of course, the ability to talk about lawful optics. To encode \mintinline{haskell}{Getter}s and \mintinline{haskell}{Review}s, we then set $\C_2$ or $\C_1$ to be the codiscrete category $\C_\mathsf{codisc}$, i.e. the category with the same set of objects as $\C$ but with all homsets the singleton set. In this way we indeed get a category $\Getter$ that is isomorphic to $\C$:
\begin{align*}
\int^{M \in \M} \C(S, M \times A) \times \C_\mathsf{codisc}(M \times A', S')
&\cong \int^{M \in \M} \C(S, M \times A) \\
&\cong \C(S, 1 \times A) \\
&\cong \C(S, A)
\end{align*}
This works for $\Review$ in the same way, putting $\C_\mathsf{codisc}$ on the left. The functor $\C \to \C_\mathsf{codisc}$ induces functors $\Lens \to \Getter$ and $\Prism \to \Review$.

Finally, we can also encode \mintinline{haskell}{Fold}s, which in $\Set$ correspond to functions $S \to \mathrm{List}\,A$. The key is that the $\mathrm{List}$ functor is the terminal $\Traversable$, so we have isomorphisms
\begin{align*}
&\int^{M \in \Traversable} \Set(S, MA) \times \Set_\mathsf{codisc}(MA', S') \\
&\cong \int^{M \in \Traversable} \Set(S, MA) \\
&\cong \Set(S, \mathrm{List}\,A)
\end{align*}

@@

\begin{itemize}
\item \emph{``Achromatic'' Lenses}~\cite[Section 5.2]{ProfunctorOpticsThesis} are lenses that also admit an operation $\fcreate : A \to S$. These are optics for the action of $\C$ on itself by $M \act A = (M \sqcup 1) \times A$, or equivalently, of the category of pointed objects of $\C$ on $\C$ by cartesian product. Concrete achromatic lenses $S \hto A$ are elements of the set \[\C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S').\]
%\begin{align*}
%  \Optic((S, S'), (A, A'))
%  &= \int^{M \in \C} \C(S, M \act A) \times \C(M \act A', S') \\
%  &= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
%  &\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
%  &\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S')
%\end{align*}
\item \emph{Affine Traversals}~\cite{AffineTraversalPost} allow access to a target that may or may not be present. Suppose $\C$ is cartesian closed and has binary coproducts. Let $\mathsf{Aff}$ be the category $\C \times \C$, equipped with the monoidal structure
\begin{align*}
  (P', Q') \otimes (P, Q) &= (P' \sqcup (Q' \times P) , Q' \times Q)
\end{align*}
The category $\mathsf{Aff}$ acts on $\C$ by $(P, Q) \act A = P \sqcup (Q \times A)$, in fact, $\mathsf{Aff}$ is cooked up exactly as the closure of the actions $- \times A$ and $- \sqcup A$ under composition. A concrete affine traversal is an element of \[\C(S, S' \sqcup (\homC(A', S') \times A)).\]

@@

@@

>>>>

\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\const}{\mathrm{const}}
\DeclareMathOperator{\ob}{ob}
\DeclareMathOperator{\copr}{copr}
\newcommand{\inl}{\mathrm{inl}}
\newcommand{\inr}{\mathrm{inr}}
\DeclareMathOperator{\im}{im}
\newcommand{\act}{\cdot}
\newcommand{\codisc}{\mathsf{codisc}}

@@

\begin{proposition}\label{prop:iota-functor}
  There is a functor $\iota : \C \times \C^\op \to \Optic_\otimes$, which on objects is given by $\iota(S, S') = (S, S')$ and on a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \rep{\lambda_A^{-1} f}{g \lambda_{A'}}$.
\end{proposition}
\begin{proof}
  Graphically, this is:
  \begin{center}
    \input{diagrams/iota.tikz}
  \end{center}

@@

\begin{proposition}
  For monoidal $F : \C \to \D$, the induced $\Optic(F) : \Optic_\C \to \Optic_\D$ is also monoidal.
\end{proposition}
\begin{proof}
The structure morphisms for monoidality are given by lifting the structure morphisms for $F$:
\begin{align*}
\phi_{(S, S'), (T, T')} &:= \iota(\phi_{S, T}, \phi_{S', T'}) &&: F(S, S') \otimes F(T, T') \hto F((S, S') \otimes (T, T')) \\
\phi &:= \iota(\phi, \phi) &&: (I, I) \hto F(I, I)
\end{align*}
The axioms follow by lifting the axioms for $F$. More difficult is verifying naturality: \todo{todo!}
\end{proof}

@@

\begin{definition}
  Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category. An \emph{action of $\M$ on $\C$} is a monoidal functor $a : \M \to [\C, \C]$. For two objects $M \in \M$ and $A \in \C$, the action $a(M)(A)$ is abbreviated $M \act A$ or $MA$. The tensor product in $\M$ will also be written as juxtaposition, so $(NM)A \cong N(MA)$.
\end{definition}

@@

\begin{proposition}
  We have a category $\Optic_\M$ and a functor $\iota : \C \times \C^\op \to \Optic_\M$ defined analogously to Propositions~\ref{prop:optic-is-cat} and~\ref{prop:iota-functor}. \qed
\end{proposition}

\begin{definition}
Given two categories equipped with monoidal actions $(\M, \C)$ and $(\N, \D)$, a \emph{morphism of actions} is a monoidal functor $F^\bullet : \M \to N$ and a functor $F : \C \to \D$ that commutes with the actions, in the sense that there exist natural isomorphisms
  \begin{align*}
  \phi &: F(I A) \to (F^\bullet I)(FA) \\
  \phi_{M,A} &: F(MA) \to (F^\bullet M)(F A)
  \end{align*}
satisfying conditions analogous to those for a monoidal functor.

%A morphism of actions is \emph{lax} if $\phi$ and $\phi_{M,A}$ are not necessarily invertible, and \emph{oplax} if they face the other direction. \todo{Do I need laxness in $F^\bullet$?}
\end{definition}

\begin{proposition}\label{prop:change-of-action}
If $F : (\M, \C) \to (\N, \D)$, is a morphism of actions, there is an induced functor $\Optic(F) : \Optic_\M \to \Optic_\N$. \qed
\end{proposition}

@@

\begin{proposition}\label{prop:lawful-lens-laws}
  A concrete lens given by $\fget$ and $\fput$ is lawful (in our sense) iff it obeys the three concrete lens laws.
\end{proposition}
\begin{proof}
  We begin by giving $\Twoptic_\times(S, A)$ the same treatment as $\Optic_\times(S, A)$. Using Yoneda reduction and the universal property of the product as we did for lenses themselves, we have:
  \begin{align*}
  \Twoptic_\times(S, A)
  &= \int^{M_1, M_2 \in \C} \C(S, M_1 \times A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_1, M_2 \in \C} \C(S, M_1) \times \C(S, A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2) \times \C(S \times A, A) \times \C(M_2 \times A, S) \\
  &\cong \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)
  \end{align*}
  Call this last set $\conctwice_\times(S, A)$. Written equationally, the isomorphism $\Phi : \Twoptic_\times(S, A) \to \conctwice_\times(S, A)$ is given by:
  \begin{align*}
    \Phi(\repthree{l}{c}{r}) = (&\pi_2 l, \\
                                   &\pi_2 c (\pi_1 l \times A), \\
                                   &r(\pi_1 c (\pi_1 l \times A) \times A))
  \end{align*}
  Now suppose we are given a lens $p$ that corresponds concretely to $(\fget, \fput)$, so $p = \rep{[\id_S, \fget]}{\fput}$. Evaluating $\outside$ on this gives:
  \begin{align*}
    \outside(\rep{[\id_S, \fget]}{\fput}) = \fput [\id_S, \fget]
  \end{align*}
  so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

@@

\begin{definition}
  Suppose $\C$ has finite coproducts. The \emph{category of prisms} is the category of lawful optics with respect to the coproduct $\sqcup$.
\end{definition}

@@

There is a common pattern in many of the examples to follow: for every object $A \in \C$, the evaluation-at-$A$ functor $- \act A : \M \to \C$ has a right adjoint, say $R_A : \C \to \M$. To fix notation, let \[\overrightarrow{(-)} : \C(F \act A, S) \to \M(F, R_{A} S) : \overleftarrow{(-)}\] denote the homset bijection, so the unit and counit are:
  \begin{align*}
  	\eta_F &: F \to R_A (F \act A) \\
    \eta_F &:= \overrightarrow{\id_{F \act A}} \\
    \varepsilon_{S} &: (R_A S) \cdot A \to S \\
    \varepsilon_{S} &:= \overleftarrow{\id_{R_{A} S}}
  \end{align*}

@@

In this case we can always eliminate the coend and find a concrete description of an optic:
\begin{align*}
&\int^{F \in \M} \C(S, F\act A) \times \C(F\act A', S') \\
&\cong \int^{F \in \M} \C(S, F \act A) \times \M(F, R_{A'} S') \\
&\cong \C(S,  (R_{A'} S') \act A)
\end{align*}
A concrete optic is therefore a map $\funzip : S \to (R_{A'} S') \act A$. The above isomorphism sends $\funzip$ to the element $\rep{\funzip}{\varepsilon_{S'}}$. In the other direction, given $\rep{l}{r}$ we have $\funzip = (\overrightarrow{r} \act A)l$.

@@

\begin{definition}
The category $\Traversal$ of traversals is the category of lawful optics for the action of $\Traversable$ on $\Set$ given by evaluation. (Yes, the names $\Traversal$ vs\@. $\Traversable$ are confusing!)
\end{definition}

@@

Suppose $\C$ is a monoidal closed category and $T : \C \to \C$ is a strong monad. Then the evaluation-at-$A$ functor has a right adjoint:
\begin{align*}
\C_T(M \monact A', S') 
&= \C(M \times A', T S') \\
&\cong \C(M, \homC(A', T S'))
\end{align*}
Using the coalgebraic description, we see that concrete effectful lenses consist of a single morphism \[\munzip : S \to T(\homC(A', T S') \otimes A).\] The optic laws in this case specialise to:

\begin{proposition}
A concrete effectful lens is lawful iff
  \begin{align*}
    \mu_S T(\ev_{A, TS}) \; \munzip &= \eta_S \\
    T(\eta_{\homC(A, TS) \otimes A}\coev_{\homC(A, TS), A} \otimes A)\munzip &= T((\munzip \circ_T -) \otimes A)\munzip
  \end{align*}
  where \[ \munzip \circ_T - : \homC(A, TS) \to \homC(A, T(\homC(A, TS) \otimes A)) \] denotes internal Kleisli composition and \[\coev_{\homC(A, TS), A} : \homC(A, TS) \to \homC(A, \homC(A, TS) \otimes A) \] is coevaluation. \qed
\end{proposition}
  Or, if you prefer do-notation, the two laws are:
\begin{multicols}{2}
\begin{minted}{haskell}
do (c, a) <- munzip s
   c a
==
return s
\end{minted} 
~\columnbreak
\begin{minted}{haskell}
do (c, a) <- munzip s
   let f a' = do
     s' <- c a'
     munzip s'
   return (f, a)
==
do (c, a) <- munzip s
   let f a' = (c, a')
   return (f, a)
\end{minted}
\end{multicols}

The category $\C$ includes into $\C_T$ in a way that preserves the action of $\C$, so there is an induced inclusion $\Optic_\otimes \to \Optic_\monact$.

@@

Suppose $\C$ is cartesian closed.

@@

We can find a concrete description of optics for the action $\monact : \C \times \C_{T_Q} \to \C_{T_Q}$ that is closer to that for ordinary lenses.
\begin{align*}
\Optic((S, S'), (A, A'))
&= \int^{M \in \C} \C_{T_Q}(S, M \monact A) \times \C_{T_Q}(M \monact A', S') \\
&= \int^{M \in \C} \C(S, \homC(Q, M \times A \times Q)) \times \C_{T_Q}(M \monact A', S') \\
&\cong \int^{M \in \C} \C(S \times Q, M \times A\times Q) \times \C_{T_Q}(M \monact A', S') \\
&\cong \int^{M \in \C} \C(S \times Q, M) \times \C(S \times Q, A\times Q) \times \C_{T_Q}(M \monact A', S') \\
&\cong \C(S \times Q, A \times Q) \times \C_{T_Q}((S \times Q) \monact A', S') \\
&\cong \C_{T_Q}(S, A) \times \C_{T_Q}(S \times Q \times A', S')
\end{align*}
By analogy with ordinary lenses, let us call these maps $\mget$ and $\mput$.

@@

\begin{proposition}
\todo{Translate this from code}
A stateful lens given by
\begin{minted}{haskell}
mget :: s -> State q a
mput :: s -> q -> a -> State q s
\end{minted}
is lawful iff the following three laws hold: \\
\begin{minipage}{\textwidth}
\begin{multicols}{3}
\begin{minted}{haskell}
do
  q <- getState
  a <- mget s
  mput s q a
==
return s
\end{minted}
~\columnbreak
\begin{minted}{haskell}
do s' <- mput s q a
   mget s'
==
return a
\end{minted}
~\columnbreak
\begin{minted}{haskell}
let (s', q') 
  = runState (mput s q1 a1) q2
in mput s' q' a2
==
mput s q1 a2
\end{minted}
\end{multicols}
\end{minipage}
\end{proposition}
We may as well also name these the $\fget\fput$, $\fput\fget$ and $\fput\fput$ laws.
\begin{proof}
%We similarly calculate
%\begin{align*}
%&\int^{M, N \in \C} \C_{T_Q}(S, M \monact A) \times \C_{T_Q}(M \monact A, N \monact A) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C(S, \homC(Q, M \times A \times Q)) \times \C(M \times A, \homC(Q, N \times A \times Q)) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C(S \times Q, M \times A \times Q) \times \C(M \times A \times Q, N \times A \times Q) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{M, N \in \C} \C(S \times Q, M) \times \C(S \times Q,  A \times Q) \times \C(M \times A \times Q, N \times A \times Q) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{N \in \C} \C_{T_Q}(S, A) \times \C(S \times Q \times A \times Q, N \times A \times Q) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \int^{N \in \C} \C_{T_Q}(S, A) \times \C(S \times Q \times A \times Q, N) \times \C(S \times Q \times A \times Q, A \times Q) \times \C_{T_Q}(N \monact A, S) \\
%&\cong \C_{T_Q}(S, A)\times \C_{T_Q}((S \times Q) \monact A, A) \times \C_{T_Q}((S \times Q \times A \times Q) \monact A, S)
%\end{align*}
\todo{This is a mess}
\end{proof}

@@

@@

\begin{itemize}
\item \emph{``Achromatic'' Lenses}~\cite[Section 5.2]{ProfunctorOpticsThesis} are lenses that also admit an operation $\fcreate : A \to S$. These are optics for the action of $\C$ on itself by $M \act A = (M \sqcup 1) \times A$, or equivalently, of the category of pointed objects of $\C$ on $\C$ by cartesian product. Concrete achromatic lenses $(S, S') \hto (A, A')$ are elements of the set \[\C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S').\]
%\begin{align*}
%  \Optic((S, S'), (A, A'))
%  &= \int^{M \in \C} \C(S, M \act A) \times \C(M \act A', S') \\
%  &= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
%  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
%  &\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
%  &\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S')
%\end{align*}
\item \emph{Affine Traversals}~\cite{AffineTraversalPost} allow access to a target that may or may not be present. Suppose $\C$ is cartesian closed and has binary coproducts. Let $\mathsf{Aff}$ be the category $\C \times \C$, equipped with the monoidal structure
\begin{align*}
  (P', Q') \otimes (P, Q) &= (P' \sqcup (Q' \times P) , Q' \times Q)
\end{align*}
The category $\mathsf{Aff}$ acts on $\C$ by $(P, Q) \act A = P \sqcup (Q \times A)$, in fact, $\mathsf{Aff}$ is cooked up exactly as the closure of the actions $- \times A$ and $- \sqcup A$ under composition. A concrete affine traversal is an element of \[\C(S, S' \sqcup (\homC(A', S') \times A)).\]

@@

\item \emph{Grates}~\cite{GratesPost} are optics for the contravariant action of a monoidal closed category $\C$ on itself by $X \cdot A \mapsto \homC(X, A)$. Concretely these correspond to morphisms \[ \C(\homC(\homC(S, A), A'), S'). \]

@@

\section{Mixed Optics}\label{sec:mixed-optics}

We can generalise the definition of $\Optic$ so that the two halves lie in different categories, allowing us to capture a few more optic variants found in the wild. Suppose $\C_1$ and $\C_2$ are categories that are each acted on by a common monoidal category $\M$.

\begin{definition}
  Given two objects of $\C_1 \times \C_2^\op$, say $(S, S')$ and $(A, A')$, an \emph{optic} $p : (S, S') \hto (A, A')$ is an element of the set
  \begin{align*}
    \Optic_\otimes((S, S'), (A, A')) := \int^{M \in \M} \C_1(S, M \act A) \times \C_2(M \act A', S')
  \end{align*}
\end{definition}

Several results from Section~\ref{sec:optics} continue to hold:
\begin{proposition}\label{prop:mixed-optic-is-cat}
  The above data form a category $\Optic_{\M, \C_1, \C_2}$. \todo{notation?} \qed
\end{proposition}
\begin{proposition}
  There is a functor $\iota : \C_1 \times \C_2^\op \to \Optic_{\M, \C_1, \C_2}$, which on objects is given by $\iota(S, S') = (S, S')$ and on a morphism $(f, g) : (S, S') \to (A, A')$ by \[\iota(f, g) = \rep{(\lambda^{\C_1}_A)^{-1} f}{g \lambda^{\C_2}_{A'}},\] where $\lambda^{\C_1}_A : A \to IA$ in $\C_1$ and $\lambda^{\C_2}_{A'} : A' \to IA'$ in $\C_2$. \qed
\end{proposition}
\begin{proposition}\label{prop:mixed-change-of-action}
  Given two morphisms of actions $F_1 : (\M, \C_1) \to (\N, \D_1)$ and $F_2 : (\M, \C_2) \to (\N, \D_2)$, there is an induced functor $\Optic(F_1, F_2) : \Optic_{\M, \C_1, \C_2} \to \Optic_{\N, \D_1, \D_2}$. \qed
\end{proposition}

The notion of lawfulness used earlier breaks down in this setting, as the definition of $\Twoptic(S, A)$ only makes sense when $\C_1 = \C_2$. The following definition gives a reasonable approximation.

\begin{definition}
If $F_1 : (\M, \C_1) \to (\N, \D)$ and $F_2 : (\M, \C_2) \to (\N, \D)$, a mixed optic $p : (S, S) \hto (A, A)$ in $\Optic_{\M, \C_1, \C_2}$ is \emph{lawful relative to $(F_1, F_2)$} if its image under $\Optic(F_1, F_2)$ is lawful.
\end{definition}

\subsection{Examples}

\subsubsection{Degenerate Optics}

The \mintinline{haskell}{Getter}s and \mintinline{haskell}{Review}s of the Haskell lens library corresponds simply to a morphisms $S \to A$ and $A' \to S'$ respectively. These are examples of mixed optics as follows.

Recall that for any category $\C$, there is a category $\C_\codisc$ with the same objects, but where every homset is the singleton set. For any action of $\M$ on $\C$ there is a corresponding action of $\M$ on $\C_\codisc$ that has the same action on objects, and a morphism of actions $i : (\M, \C) \to (\M, \C_\codisc)$ which is the identity on objects.
